Technique for \earlywarning\ detection of compact binary coalescence and its implication for multi-messenger astronomy

### 2011-06-03 method.tex
Edited sections 2.4 and 2.5
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@888 ccdd05d5-56c1-461a-bf98-85f6876523b4

As described previously, the template banks are, by design, highly correlated.
It is possible to greatly reduce the number of filters required to achieve a
particular minimum match by designing an appropriate set of orthonormal {\em
basis templates}.  A purely numerical technique based on the application of the
singular value decomposition (\SVD) to inspiral waveforms is demonstrated
in~\cite{Cannon:2010p10398}.  Using the results of ~\cite{Cannon:2010p10398} we
establish that the time sliced templates of the previous section can be
approximated to high accuracy by the expansion
%
\begin{equation}
h_i^s[k] = \sum_{\mathclap{l=0}}^{\mathclap{M-1}} v_{il}^s \sigma_l^s u_l^s[k] \approx \sum_{\mathclap{l=0}}^{\mathclap{L^s-1}} v_{il}^s \sigma_l^s u_l^s[k]
\label{eq:svddecomp}
\end{equation}
%
where $\sigma_l^s$ is the $l^{\mathrm{th}}$ singular value for the
$s^{\mathrm{th}}$ time slice, $v_{il}^s$ is an orthogonal matrix for the
$s^{\mathrm{th}}$ time slice, and $u_l^s[k]$ is a new set of orthogonal basis filters
for the $s^{\mathrm{th}}$ time slice.  The authors of \cite{Cannon:2010p10398}
showed that to high reconstruction accuracy far fewer filters are needed than
were in the original template bank. We find that when combined with the
time-slice decomposition, the number of \SVD\ filters, \numsvdtmps\ is much smaller
than the original number of filters \numtmps.  We combine \eqref{eq:svddecomp}
with \eqref{eq:time-slices} to arrive at \eqref{eq:decomp}.  In the next
section we compute the expected computational cost scaling of this
decomposition and compare it with the brute-force implementation in
\eqref{eq:SNRTD} and higher latency \fft\ methods.

@@

\editorial{I just dropped our new notation in here; I haven't edited the text yet.}%
Definition of decimated detector data:
\begin{equation}
	x^{s+1}[k] = \left( H^\shortdownarrow x^s\right)[k]
\end{equation}

In order to minimize latency we propose using the time-domain convolution
presented in \eqref{eq:SNRTD}.  However, because the brute-force evaluation of
\eqref{eq:SNRTD} is far too costly to be useful, we will consider an
approximation to \eqref{eq:SNRTD} that can reduce substantially the cost of
\realtime\ filtering. This approximation has the form
%
% orthogonal decomposition filtering
%
\begin{equation}
	\rho_i^s [k] =%
		% Reconstruction
		\underbrace{
			\sum_{\mathclap{l=0}}^{\mathclap{L^s-1}} v_{il}^s \sigma_l^s
		}_\textrm{\clap{reconstruction}}
		% Orthogonal FIR filter
		\overbrace{
			\sum_{\mathclap{n=0}}^{\mathclap{N^s-1}} u_l^s[n] x^s[k-n]
		}^\textrm{\clap{orthogonal {\sc fir} filters}}
		% Plus ...
		+
		% Interpolation SNR
		\underbrace{
			\left(H^\uparrow \rho_i^{s+1}\right)[k]
		}_\textrm{\clap{{\sc snr} from previous time slices}}
\end{equation}
%
%
where $u_{jk}(\tau)$ is an orthogonal basis set of filters spanning the space
of $\{x_i(\tau)\}$ and $\sigma_{jk} v_{ijk}$ is a tensor relating the filters
$u_{jk}(\tau)$ to the original filter set $\{x_i(\tau)\}$.  We claim that with
a suitable choice of filters $u_{jk}(\tau)$ one can reduce the computational
cost of \eqref{eq:SNRTD} sufficiently to feasibly search for
gravitational waves from compact binary coalescence in \realtime.  This
requires 1) exploiting
the time-frequency characteristics of the binary waveforms and 2) exploiting the redundancy of the template bank. We describe our
procedure for producing the decompostion in \eqref{eq:decomp} in the remainder
of this section.

@@

We now examine the computational cost scaling of the approximate implementation
of a conventional time domain or frequency domain matched filter bank as compared
with \lloid{}.  For convenience, \ref{tab:recap} provides a review of the notation
that we will need in this section.
%
% symbols used in FLOPs calculations table
%
\Table{\label{tab:recap} Notation used to describe filters.}
\br
Symbol	& Definition \\
\mr
\numtmps		& The number of templates \\
\tmpsamps	& The number of samples per template \\
\numslices	& The number of time slices \\
\numsvdtmps	& The number of orthogonal templates in time slice $s$ \\
\slicessamps	& The number of samples in time slice $s$ \\
$f^s$		& The sample rate in time slice $s$ \\
$N^\shortdownarrow$ & The number of coefficients in the decimation filter \\
$N^\shortuparrow$ & The number of coefficients in the interpolation filter \\
\br
\end{tabular}
\end{indented}
\end{table}

>>>>

As described previously, the template banks are, by design, highly correlated.
It is possible to greatly reduce the number of filters required to achieve a
particular minimum match by designing an appropriate set of orthonormal {\em
basis templates}.  A numerical technique based on the
singular value decomposition (\SVD) of inspiral template banks is demonstrated
in~\cite{Cannon:2010p10398}.  Similarly, the time sliced templates described above
can be approximated to arbitrary accuracy by expansion into a set of orthonormal
\emph{basis templates}, $u_l^s[k]$, related to the original time sliced templates
through the \emph{reconstruction matrix}, $v_{il}^s\sigma_l^s$:
%
\begin{equation}
h_i^s[k] = \sum_{\mathclap{l=0}}^{\mathclap{M-1}} v_{il}^s \sigma_l^s u_l^s[k] \approx \sum_{\mathclap{l=0}}^{\mathclap{L^s-1}} v_{il}^s \sigma_l^s u_l^s[k].
\label{eq:svddecomp}
\end{equation}
%
\editorial{Say a word or two about \SVD\ tolerance?}%
%
The parameter $\numsvdtmps$ sets the number of basis templates that are kept in the
approximation.  This determines the \SVD\ tolerance, which affects the \SNR\ loss due
to the approximation.  The authors of \cite{Cannon:2010p10398} showed that high accuracy
could be achieved with far fewer basis templates than templates in the original template
bank.  We find that when combined with the time slice decomposition, the number of basis 
templates \numsvdtmps\ is much smaller than the original number of templates \numtmps.
In the next section we describe how we form our novel detection statistic using the time slice decomposition and the \SVD.

@@

Definition of decimated detector data:
\begin{equation}
	x^{s+1}[k] = \left( H^\shortdownarrow x^s\right)[k]
\end{equation}

\editorial{Chad: This whole subsection needs a lot of work.
I just dropped our new notation in here.}%
In order to minimize latency we propose using the time-domain convolution
presented in \eqref{eq:SNRTD}.  However, because the brute-force evaluation of
\eqref{eq:SNRTD} is far too costly to be useful, we will consider an
approximation to \eqref{eq:SNRTD} that can reduce substantially the cost of
\realtime\ filtering. This approximation has the form
%
% orthogonal decomposition filtering
%
\begin{equation}
	\rho_i^s [k] =%
		% Reconstruction
		\underbrace{
			\sum_{\mathclap{l=0}}^{\mathclap{L^s-1}} v_{il}^s \sigma_l^s
		}_\textrm{\clap{reconstruction}}
		% Orthogonal FIR filter
		\overbrace{
			\sum_{\mathclap{n=0}}^{\mathclap{N^s-1}} u_l^s[n] x^s[k-n]
		}^\textrm{\clap{orthogonal {\sc fir} filters}}
		% Plus ...
		+
		% Interpolation SNR
		\underbrace{
			\left(H^\uparrow \rho_i^{s+1}\right)[k]
		}_\textrm{\clap{{\sc snr} from previous time slices}}
\end{equation}
%
%
where $u_{jk}(\tau)$ is an orthogonal basis set of filters spanning the space
of $\{x_i(\tau)\}$ and $\sigma_{jk} v_{ijk}$ is a tensor relating the filters
$u_{jk}(\tau)$ to the original filter set $\{x_i(\tau)\}$.  We claim that with
a suitable choice of filters $u_{jk}(\tau)$ one can reduce the computational
cost of \eqref{eq:SNRTD} sufficiently to feasibly search for
gravitational waves from compact binary coalescence in \realtime.  This
requires 1) exploiting
the time-frequency characteristics of the binary waveforms and 2) exploiting the redundancy of the template bank. We describe our
procedure for producing the decompostion in \eqref{eq:decomp} in the remainder
of this section.

In the next section we compute the expected computational cost scaling of this
decomposition and compare it with the brute force time domain implementation of
\eqref{eq:SNRTD} and higher latency frequency domain methods.

@@

We now examine the computational cost scaling of the approximate implementation
of a conventional time domain or frequency domain matched filter bank as compared
with \lloid{}.  For convenience, table~\ref{tab:recap} provides a review of the notation
that we will need in this section.
%
% symbols used in FLOPs calculations table
%
\Table{\label{tab:recap} Notation used to describe filters.}
\br
Symbol	& Definition \\
\mr
\numtmps		& The number of templates \\
\tmpsamps	& The number of samples per template \\
\numslices	& The number of time slices \\
\numsvdtmps	& The number of orthogonal templates in time slice $s$ \\
\slicessamps	& The number of samples in time slice $s$ \\
$f^s$		& The sample rate in time slice $s$ \\
$N^\shortdownarrow$ & The number of coefficients in the decimation filter \\
$N^\shortuparrow$ & The number of coefficients in the interpolation filter \\
\br
\end{tabular}
\end{indented}
\end{table}
